---
title: "DTAG lunges"
output:
  html_document:
    df_print: paged
---

Lunge rates of tagged whales by species, region, and prey type from DTAG (plus a couple Acousondes) data.

```{r setup, echo = FALSE, message = FALSE}
library(datetimeutils)
library(ggpubr)
library(kableExtra)
library(knitr)
library(lubridate)
library(lutz)
library(oce)
library(R.matlab)
library(readxl)
library(RcppRoll)
# Using the dev version of tidyr to use unnest_wider and unnest_longer
library(tidyverse)
library(zeallot)

opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
opts_chunk$set(echo = FALSE, message = FALSE, cache = TRUE)
```

# Tag guide summary

```{r util}
# Convert MATLAB DN's to POSIXct
dn_to_posix <- function(dn) {
  as.POSIXct((dn - 719529) * 86400, origin = "1970-01-01", tz = "UTC")
}

# Regex patterns for tag-type IDs
cats_ID <- "[a-z]{2}[0-9]{6}-[0-9]{1,2}[a-z]?"
dtag_ID <- "[a-z]{2}[0-9]{2}_[0-9]{1,3}[a-z]"
acou_ID <- "[a-z]{2}[0-9]{6}[a-z]?"
parse_tagtype <- function(id) {
  case_when(str_detect(id, cats_ID) ~ "CATS",
            str_detect(id, dtag_ID) ~ "DTAG",
            str_detect(id, acou_ID) ~ "ACOU",
            TRUE ~ "UNKTAG")
}

POSIX_from_excel <- function(datenum) {
  convert_date(parse_number(datenum), "excel", fraction = TRUE, tz = "UTC")
}

find_prh <- function(id, tag) {
  prh_dir <- case_when(tag == "ACOU" ~ "/Volumes/COPYCATSdat/ACOU/prh",
                       tag == "DTAG" ~ "/Volumes/COPYCATSdat/DTAG/prh",
                       TRUE ~ "~")
  map2_chr(prh_dir, id, function(.x, .y) {
    result <- dir(.x, pattern = .y, full.names = TRUE)
    if (length(result) == 0) NA_character_
    else if (length(result) == 1) result
    else "MULTIPLE"
  })
}

find_lunge <- function(id, tag) {
  lunge_dir <- case_when(tag == "ACOU" ~ "/Volumes/COPYCATSdat/ACOU/lunges",
                         tag == "DTAG" ~ "/Volumes/COPYCATSdat/DTAG/lunges",
                         TRUE ~ "~")
  map2_chr(lunge_dir, id, function(.x, .y) {
    result <- dir(.x, pattern = .y, full.names = TRUE)
    if (length(result) == 0) NA_character_
    else if (length(result) == 1) result
    else "MULTIPLE"
  })
}

find_tz <- function(lon, lat) {
  tz_lookup_coords(lat, lon, "fast", FALSE)
}

force_tzs2 <- function(time, tzones, tzone_out = "UTC") {
  result_list <- map2(time, tzones, function(.x, .y) {
    tryCatch(force_tzs(.x, .y, tzone_out), error = function(e) NA)
  })
  result_vec <- unlist(result_list)
  attributes(result_vec) <- attributes(result_list[[1]])
  result_vec
}
```

```{r solar}
sun_phase <- function(x, y, t) {
  alt <- pmap_dbl(list(x, y, t), function(x, y, t) {
    if (any(is.na(c(x, y, t)))) {
      NA
    } else {
      alt <- sunAngle(t, longitude = x, latitude = y)$altitude
    }
  })
  cut(alt, c(-Inf, -18, 0, Inf), labels = c("NIGHT", "TWILIGHT", "DAY"))
}

daylight_fun <- function(start, end, lon, lat, n = 1e3) {
  pmap_dbl(list(start, end, lon, lat), function(start, end, lon, lat) {
    if (any(is.na(c(start, end, lon, lat)))) {
      NA
    } else {
      angle <- sunAngle(seq(start, end, length.out = n), lon, lat)$altitude 
      as.numeric(end - start, units = "hours") * sum(angle > 0) / n
    }
  })
}
twilight_fun <- function(start, end, lon, lat, n = 1e3) {
  pmap_dbl(list(start, end, lon, lat), function(start, end, lon, lat) {
    if (any(is.na(c(start, end, lon, lat)))) {
      NA
    } else {
      angle <- sunAngle(seq(start, end, length.out = n), lon, lat)$altitude 
      as.numeric(end - start, units = "hours") * sum(angle >= -18 & angle <= 0) / n
    }
  })
}
night_fun <- function(start, end, lon, lat, n = 1e3) {
  pmap_dbl(list(start, end, lon, lat), function(start, end, lon, lat) {
    if (any(is.na(c(start, end, lon, lat)))) {
      NA
    } else {
      angle <- sunAngle(seq(start, end, length.out = n), lon, lat)$altitude 
      as.numeric(end - start, units = "hours") * sum(angle < -18) / n
    }
  })
}
```

```{r deployments}
deployments <- read_xlsx("data/DTag data_fixesforMFC.xlsx") %>% 
  transmute(ID, 
            species = str_sub(ID, 1, 2), 
            longitude = Long,
            latitude = Lat,
            sonar_exp = case_when(sonar_exp %in% c("yes", "PRN") ~ TRUE,
                                  sonar_exp %in% c("none", "NA") ~ FALSE),
            prh_exists = case_when(`prh file found` == "Y" ~ TRUE,
                                   is.na(`prh file found`) ~ FALSE),
            lunge_exists = case_when(`lunge file found` == "Y" ~ TRUE,
                                     is.na(`lunge file found`) ~ FALSE,
                                     `lunge file found` == "N" ~ FALSE),
            tag_type = parse_tagtype(ID),
            prh_path = find_prh(ID, tag_type),
            lunge_path = find_lunge(ID, tag_type),
            tag_tz = find_tz(longitude, latitude),
            tagon = force_tzs2(POSIX_from_excel(`Tag on`), tag_tz),
            tagoff = force_tzs2(POSIX_from_excel(`Tag off`), tag_tz),
            duration_h = as.numeric(tagoff - tagon, unit = "hours"), 
            daylight_h = daylight_fun(tagon, tagoff, longitude, latitude),
            night_h = night_fun(tagon, tagoff, longitude, latitude),
            twilight_h = twilight_fun(tagon, tagoff, longitude, latitude))
```

We have two minkes with an indecipherable tag type. There's a total of 26 fins. 23/24 fin DTAGs have PRHs and lunges. Data for the two Acousonde fins are on the shared drive and should be added (several days of data). Probably a similar story for the six blue whale Acousondes. 67 blue whale DTAGs!

```{r deployment_summ}
deployments %>% 
  group_by(species, tag_type) %>% 
  summarize(N = n(),
            Nprh = sum(prh_exists),
            Nlunge = sum(lunge_exists))
```

```{r lunge_times}
lunge_times <- function(prh_path, lunge_path) {
  if (any(is.na(c(prh_path, lunge_path))))
    return(NA)
  lunge_mat <- readMat(lunge_path)
  # If dates aren't in the lunge file, use PRH to look up by index
  if (is.null(lunge_mat$LungeDN)) {
    prh_mat <- readMat(prh_path)
    dn <- prh_mat$DN[lunge_mat$LungeI]
  } else {
    dn <- lunge_mat$LungeDN
  }
  dn_to_posix(dn)
}

lunges <- deployments %>%
  mutate(lunge_time = map2(prh_path, lunge_path, lunge_times)) %>% 
  unnest(lunge_time) %>%
  mutate(lunge_time = force_tzs2(lunge_time, tag_tz),
         lunge_sun = sun_phase(x = longitude, y = latitude, t = lunge_time)) %>% 
  select(ID, species, lunge_time, lunge_sun) %>% 
  drop_na()

save(lunges, file = "data/outputs/DTAG_lunges/lunges.RData")
```

```{r rates}
lunges_deploy <- lunges %>% 
  group_by(ID) %>% 
  summarize(total_lunges = n(),
            day_lunges = sum(lunge_sun == "DAY"),
            twilight_lunges = sum(lunge_sun == "TWILIGHT"),
            night_lunges = sum(lunge_sun == "NIGHT")) %>% 
  left_join(select(deployments, ID, species,
                   total_hours = duration_h,
                   day_hours = daylight_h,
                   night_hours = night_h),
            by = "ID")
```

```{r qa_lungerates}
lunge_rate_by_solar <- function(rows, keys) {
  c(p, rates) %<-% tryCatch({
    prh <- readMat(rows$prh_path[1])
    if (is.null(prh$DN)) {
      DN <- first(rows$tagon) + 1:nrow(prh$A) / prh$fs[1]
    } else {
      DN <- force_tz(dn_to_posix(prh$DN), rows$tag_tz[1]) %>% 
        with_tz("UTC")
    }
    speed_path <- dir('data/speeds', keys$ID, full.names = TRUE)
    if (length(speed_path) == 1) {
      speed <- speed_path %>% 
      readLines %>% 
      parse_number %>% 
      roll_mean(n = 10, fill = NA)
    } else {
      speed <- rep(NA, length(DN))
    }
    lon <- mean(rows$longitude)
    lat <- mean(rows$latitude)
    n_hours <- 6
    prh_data <- tibble(datetime = with_tz(DN, rows$tag_tz[1]),
                       depth = prh$p,
                       speed) %>% 
      slice(seq(1, length(DN), length.out = 1e4)) %>% 
      mutate(sunalt = sunAngle(datetime, lon, lat)$altitude)
    rows2 <- mutate(rows, 
                    datetime = lunge_time,
                    i = findInterval(lunge_time, prh_data$datetime),
                    depth = prh_data$depth[i],
                    speed = prh_data$speed[i])
    solar <- prh_data %>% 
      mutate(phase = case_when(sunalt < -18 ~ "NIGHT",
                               sunalt < 0   ~ "TWILIGHT",
                               sunalt >= 0  ~ "DAY") %>% 
               factor(levels = c("NIGHT", "TWILIGHT", "DAY")),
             phaseid = cumsum(phase != lag(phase, default = first(phase)))) %>% 
      group_by(phase, phaseid) %>% 
      summarize(start = first(datetime)) %>% 
      ungroup %>% 
      arrange(phaseid) %>% 
      mutate(end = lead(start, default = last(prh_data$datetime)))
    solarpal <- c(DAY = rgb(86, 180, 233, maxColorValue = 255), # sky blue
                  TWILIGHT = rgb(213, 94, 0, maxColorValue = 255), # vermillion
                  NIGHT = rgb(0, 158, 115, maxColorValue = 255)) # bluish green
    depth_plot <- ggplot(prh_data, aes(datetime, depth)) +
      geom_rect(aes(xmin = start, 
                    xmax = end, 
                    ymin = min(prh_data$depth, na.rm = TRUE),
                    ymax = max(prh_data$depth, na.rm = TRUE),
                    fill = phase),
                data = solar,
                alpha = 0.5,
                inherit.aes = FALSE) +
      geom_line(size = 0.1) +
      geom_point(data = rows2, color = "red") +
      scale_y_reverse() +
      scale_fill_manual(values = solarpal) +
      labs(title = keys$ID,
           y = "Depth (m)") +
      theme_classic() +
      theme(legend.position = "none",
            axis.title.x = element_blank(),
            axis.text.x = element_blank())
    speed_plot <- ggplot(prh_data, aes(datetime, speed)) +
      geom_line(size = 0.1) +
      geom_point(data = rows2, color = "red") +
      labs(x = "Local time",
           y = "Speed (m/s)") +
      theme_classic() +
      theme(legend.position = "none")
    lunging <- summarize(rows, 
                         l = n(),
                         ld = sum(lunge_sun == "DAY"),
                         ln = sum(lunge_sun == "NIGHT"),
                         lt = sum(lunge_sun == "TWILIGHT")) %>% 
      gather(lvar, Lunges, l:lt)
    durs <- tibble(t = rows$duration_h[1],
                   td = rows$daylight_h[1],
                   tn = rows$night_h[1],
                   tt = rows$twilight_h[1]) %>% 
      gather(tvar, Time, t:tt)
    rates <- cbind(lunging, durs) %>% 
      mutate(Rate = Lunges / Time,
             Phase = c("Total", "Day", "Night", "Twilight")) %>% 
      select(Phase, Lunges, Time_h = Time, Rate)
    p <- ggarrange(depth_plot, 
                   speed_plot,
                   ggtexttable(rates, rows = NULL, theme = ttheme("blank")),
                   nrow = 3, ncol = 1, align = "v",
                   heights = c(3, 3, 2))
    list(p, rates)
  }, error = function(e) {
    p <- ggplot(data.frame(x = 1, y = 1), aes(x, y)) + 
      geom_point()
    rates <- tibble(Phase = numeric(0), Lunges = numeric(0), Time_h = numeric(0), Rate = numeric(0))
    list(p, rates)
  })
  ggsave(sprintf("figs/qa/lunge_rates_DTAG/%s.pdf", keys$ID),
         p,
         height = 11,
         width = 8.5)
  rates
}
lunge_rates <- lunges %>% 
  left_join(select(deployments, ID, prh_path, duration_h, daylight_h, night_h,
                   twilight_h, tag_tz, longitude, latitude, tagon, tagoff), by = "ID") %>% 
  group_by(ID) %>% 
  group_modify(lunge_rate_by_solar) %>% 
  pivot_wider(names_from = Phase, 
              values_from = c("Lunges", "Time_h", "Rate"))

save(lunge_rates, lunge_rate_summ, weighted.var, file = "data/outputs/CATS_lunges/rates.RData")
```

